---
title: "Analysis of gender pay gap of newly-graduated maths/stats students across different percentiles."
author: "Owen F. O'Connor"
date: "2025-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, warning = FALSE, message = FALSE}
library(tidyverse)
```

## Introduction

This analysis focuses on graduates with a Level 8 degree in Natural Sciences, Mathematics, and Statistics, examining the pay gap across 3 different pay groupings.
```{r universal adjustments, warning = FALSE, message = FALSE, results = 'hide'}
library(tidyverse)
load("../gradMS.Rdata")

# Data cleaning / pre-processing 
# gradMS[,1] # statistic
# gradMS[,2] # year
# gradMS[,3] # gender
# gradMS[,4] # field of study
# gradMS[,5] # NFQ Level
# gradMS[,6] # years since graduation
# gradMS[,7] # measure of statistic

# field of study and NFQ are irrelevant since they're always the same. We can remove them
# gradMS <- gradMS[,-c(4,5)]
# gradMS[,1] # statistic
# gradMS[,2] # year
# gradMS[,3] # gender
# gradMS[,4] # years since graduation
# gradMS[,5] # measure of statistic

gradMS %>%
  filter(Gender == "Other gender")
# there only exists one observations
# we can't make any meaningful analysis off of one point.
# We are hence deciding to omit this data

gradMS <- gradMS %>%
  filter(Gender != "Other gender") # this removes 300 rows

# we think it'd also make sense to pivot this table wider
# Increases readability and will make ggplots easier

gradMS_wider <- gradMS |> # we use a base pipe here because magrittr pipes play weird with '.'
  pivot_wider(names_from = Statistic, values_from = .)

# Some of these names are awkward, so we'll simplify them.
gradMS_wider <- gradMS_wider %>%
  rename(
    p25 = `P25 Earnings of Graduates`,
    p50 = `P50 Earnings of Graduates`,
    p75 = `P75 Earnings of Graduates`
  )
```

```{r question-specific adjustments}
gradMS_Q1 <- gradMS_wider %>%
  filter(Gender != "All genders") %>%
  filter(Years.Since.Graduation == 1) %>%
  select(!Years.Since.Graduation) %>%
  pivot_wider(names_from = Gender, values_from = c(p25, p50, p75)) %>%
  mutate(
    Pay.dif.p25 = (p25_Male / p25_Female) - 1,
    Pay.dif.p50 = (p50_Male / p50_Female) - 1,
    Pay.dif.p75 = (p75_Male / p75_Female) - 1
  )

gradMS_Q1_final <- gradMS_Q1 %>%
  pivot_longer(
    cols = c(Pay.dif.p25, Pay.dif.p50, Pay.dif.p75),
    names_to = "Percentile",
    values_to = "Pay_Ratio"
  ) %>%
  # Create factor with levels in desired order (p75 at top, p25 at bottom)
  mutate(Percentile = factor(Percentile, 
                             levels = c("Pay.dif.p75", "Pay.dif.p50", "Pay.dif.p25"),
                             labels = c("75th Percentile", "50th Percentile", "25th Percentile"))) %>%
  select(Graduation.Year, Percentile, Pay_Ratio)
```

## Including Plots


```{r myplot}
gradMS_Q1_final %>%
  ggplot(aes(x = Graduation.Year, y = Pay_Ratio, fill = Pay_Ratio > 0)) +
  geom_bar(stat = "identity") +
  facet_grid(Percentile ~ ., switch = "y") +
  scale_fill_manual(values = c("#E7298AFF", "#7570B3FF"), 
                    name = "Bias", 
                    labels = c("Female", "Male")) +
  theme_minimal() +
  labs(title = "New grad pay diff @ 25th/50th/75th percentile earnings",
       subtitle = "For new maths/stats graduates",
       y = "Percentiles",
       x = "Graduation Year")
```

This plot shows clearly how the pay gap has evolved over time for maths/stats new grads across different salary percentiles.  Just looking at the first two years you one can observe how there was no difference at the lower-end of the payscale, there is a clear bias towards males at the median pay, and this is even more pronounced at the 75th percentile.  Then from this point it appears to flip, where the greater gap appears at the 25th percentile, before flipping altogether. around 10xx we can see how the market shows a greater salary bias towards females. This is pronounced at the 25th percentile, lesser at the median, and reflected in no bias at the upper percentile. This trend is short-lived though as the bias quickly returns to males and records particularly strong in 20xx.