---
title: "Group A - Data Analysis"
author: "OWEN O’CONNOR | WARREN LANGRIDGE | DAVID SKERRITT | OISIN HEANEY"
date: "2025"
output:
  html_document:
    theme: cerulean
    highlight: tango
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Packages, results='hide', warning=FALSE, message = FALSE}
library(tidyverse)
library(paletteer)
load("gradMS.Rdata")
```

# Introduction

A load of words here about into.  


# Data adjustments

A load of words here about adjustments.  
```{r adjustments}
# Data cleaning / pre-processing 
#gradMS[,1] # statistic
#gradMS[,2] # year
#gradMS[,3] # gender
#gradMS[,4] # field of study
#gradMS[,5] # NFQ Level
#gradMS[,6] # years since graduation
#gradMS[,7] # measure of statistic

# field of study and NFQ are irrelevant since they're always the same. We can remove them
gradMS <- gradMS[,-c(4,5)]
#gradMS[,1] # statistic
#gradMS[,2] # year
#gradMS[,3] # gender
#gradMS[,4] # years since graduation
#gradMS[,5] # measure of statistic

gradMS <- gradMS %>%
  filter(Gender != "Other gender") # this removes 300 rows
# *there only exists one observations
# we can't make any meaningful analysis off of one point.
# I am hence deciding to omit this data


```


# Analysis

A load of words here about analysis.

## Question 1 - Owen
'Question title'

```{r Plot 3 - OWEN O’CONNOR}
ggplot()
```

Owen's thoughts on this... 

-----

## Question 2 - Warren
'Question title'

```{r Plot 2 - WARREN LANGRIDGE}
ggplot()
```

Warren's thoughts on this...  

-----

## Question 3 - Oisin
### 'Is the despair of pay between genders closing?'  
- An analysis of pay by gender in Salary in Natural Sciences, Mathematics and Statistics for level 8 graduates
```{r Plot 1 - OISIN HEANEY}
#gender_pay_gap <- gradMS_gender %>% 
#pivot_wider(names_from = Gender, values_from = Earnings) %>% 
#mutate(Gender_Pay_Gap = (Male - Female) / Male * 100) 

#ggplot(gender_pay_gap, aes(x = Graduation.Year, y = Gender_Pay_Gap)) + 
#geom_point(color = "red", alpha = 0.8) + 
#geom_smooth(method = "lm", se = FALSE, colour = "black") + 
#labs(title = "Gender Pay Gap Over Time", 
#subtitle = " percentage difference in earnings" , 
#x = "Graduation Year", 
#y = "Gender Pay Gap (%)") + 
#theme_minimal() 
ggplot()
```

This one is a bit unclear but essentially I gathered that if the trend line goes below 0 women 
are making more than men, this makes sense as from messing around with the data it 
seems men have always generally earned more than women however the difference has 
been decreasing.   

-----

## Question 4 - David
'Question title'

```{r Plot 4 - DAVID SKERRITT}
ggplot()
```

David's thoughts on this... 

-----

# Summary

A load of words here about summary.  

